# Exam

The dataset `relinc.csv` presents the count of individuals from different religious groups according to different income ranges in the USA. The original file is structured so that each row represents a different religion, each column represents a different income range, and the cells contain the count of individuals belonging to the corresponding religion and income range.

```{r warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
relinc <- read_csv("data/relinc.csv")

relinc |> kableExtra::kable(format = "html")
```

### (a)

Read the `relinc.csv` file available in BlackBoard
<!-- [here](https://github.com/magnotairone/intro_r_python_en/blob/main/data/gapminder.csv)  -->
and adjust it to fit the tidy data format, meaning:

1.  Each variable is a column; each column represents a variable,
2.  Each observation is a row; each row represents an observation,
3.  Each value is a cell; each cell contains a single value.

The final table shoud have three columns: `Religion`, `Income`, `Count`.  

```{r include=FALSE}

relinc_tidy <- relinc |> 
  pivot_longer(cols = (ends_with("k") | "refused"), names_to = "income", values_to = "count")

relinc_tidy |> slice_sample(n=5)
```

### (b)

Return a table showing the total number of people for each religion and another table showing the total number of people in each income range.

### (c)

List the existing categories of the `religion` variable and print them. 
Select only those religions that do not start with "Other" and exclude those who refused to disclose their religion (category "refused") or income range.

### (d)

For the table resulting from item (c), create a new variable with the midpoint of each income range as follows:

| Income Range | Midpoint |
|:------------:|:--------:|
|    \<10k     |    5k    |
|    10-20k    |   15k    |
|    20-30k    |   25k    |
|    30-40k    |   35k    |
|    40-50k    |   45k    |
|    50-75k    |  62.5k   |
|   75-100k    |  87.5k   |
|   100-150k   |   125k   |
|    \>150k    | 175k \*  |

Then, calculate an approximation of the average income for each religion, using the midpoint of each income range.

### (e)

Return the religion that had, on average, the highest and lowest income.
